---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# stamp

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/randrescastaneda/stamp/branch/master/graph/badge.svg)](https://app.codecov.io/gh/randrescastaneda/stamp?branch=master)
<!-- badges: end -->

Lightweight versioned artifact store for R with sidecar metadata, pruning policies,
and Hive-style partitions.


## Installation

You can install the development version of stamp from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("randrescastaneda/stamp")
```

## Quickstart

```{r}
library(stamp)
root <- tempdir()
st_init(root)

p <- fs::path(root, "demo.qs")
x <- data.frame(id = 1:3, val = letters[1:3])

# Save with primary key (code parameter tracks provenance - see vignettes)
st_save(x, p, pk = "id")
y <- st_load(p)
vrs <- st_versions(p)
head(vrs)

# Retention
st_opts(retain_versions = 2)
st_save(transform(x, val = toupper(val)), p)
vrs <- st_versions(p)
head(vrs)

# Partitions
base <- fs::path(root, "inputs/country_year")
st_save_part(
  data.frame(country = "PER", year = 2023, pop = 34.5),
  base,
  key = list(country = "PER", year = 2023),
  pk = c("country", "year")
)
st_list_parts(base)
st_load_parts(base, as = "rbind")
```

## Managing Multiple Stamp Folders with Aliases

See the vignette "Using Aliases with stamp" for a comprehensive guide:

- Online: https://randrescastaneda.github.io/stamp/articles/using-alias.html
- Source: `vignettes/using-alias.Rmd`

## File Formats

`stamp` supports multiple serialization formats. The two binary formats have distinct implementations:

| Extension | Format | Package Required | Notes |
|-----------|--------|------------------|-------|
| `.qs2` | qs2 | `{qs2}` | New qs2 binary format (recommended for new projects) |
| `.qs` | qs | `{qs}` | Legacy qs binary format (deprecated; use `.qs2` instead) |
| `.rds` | rds | (base R) | R serialized format |
| `.csv` | csv | `{data.table}` | Comma-separated values |
| `.fst` | fst | `{fst}` | Fast columnar format |
| `.json` | json | `{jsonlite}` | JSON format |

**Important**: `.qs` and `.qs2` are **different formats** and require their respective packages. There is no automatic fallback between them. If you attempt to save/load a `.qs2` file without `{qs2}` installed, `stamp` will abort with a clear error message.

> **Migration Note**: Existing `.qs` files can still be loaded if `{qs}` is installed. To migrate to `.qs2`, load with `st_load()` and re-save with `format = "qs2"`. The `.qs2` format offers better performance and is actively maintained.

### Installing Format Packages

```r
# For qs2 format support
install.packages("qs2")

# For fst format support
install.packages("fst")
```

### Format Selection

`stamp` infers format from file extension by default:

```r
# Uses qs2 format (requires {qs2})
st_save(data, "output.qs2")

# Uses RDS (base R, always available)
st_save(data, "output.rds")
```

You can also specify format explicitly:

```r
st_save(data, "output", format = "qs2")
```

## Core Functions

The functions below are organized by workflow. **New users** should start with *Initialization*, *Save & Load*, and *Versioning* sections. Advanced features like partitions, lineage tracking, and aliases are covered in the [vignettes](#learn-more).

### Initialization & Configuration

- **`st_init(root)`** - Initialize stamp in a directory, creating the `.stamp/` state folder
- **`st_opts()`** - Get or set package options (versioning mode, retention policies, metadata format)
- **`st_opts_reset()`** - Reset all options to defaults

### Save & Load

- **`st_save(x, path, ...)`** - Save an artifact with automatic versioning, metadata, and lineage tracking
  - Optional: `pk` (primary key), `parents` (lineage), `code` (provenance), `domain` (category), `alias` (target directory)
- **`st_load(path, ...)`** - Load the latest version of an artifact
  - Optional: `verify = TRUE` (check content hash), `alias` (source directory)
- **`st_load_version(path, version_id)`** - Load a specific historical version by ID

### Versioning & History

- **`st_versions(path)`** - List all versions of an artifact with metadata (timestamp, size, hashes)
- **`st_latest(path)`** - Get the version ID of the most recent version
- **`st_changed(x, path)`** - Check if an object differs from the saved version
- **`st_changed_reason(x, path)`** - Explain why content/code changed
- **`st_should_save(x, path)`** - Determine whether saving would create a new version

### Lineage & Dependencies

- **`st_lineage(path, depth = 1)`** - Show parent artifacts (inputs) for a given artifact
- **`st_children(path, depth = 1)`** - Show child artifacts (outputs) that depend on this artifact
- **`st_is_stale(path)`** - Check if an artifact needs rebuilding because parents changed

### Metadata & Inspection

- **`st_info(path)`** - Get comprehensive artifact information (sidecar, catalog, snapshot location, parents)
- **`st_read_sidecar(path)`** - Read sidecar metadata (hashes, timestamps, primary keys, domain, parents)
- **`st_hash_obj(x)`** - Compute stable hash for any R object
- **`st_hash_code(code)`** - Compute hash of code/function
- **`st_hash_file(path)`** - Compute SHA-256 hash of file on disk

### Primary Keys

- **`st_add_pk(path, keys)`** - Add or update primary key definition for an artifact
- **`st_get_pk(x_or_meta)`** - Retrieve primary key columns from metadata
- **`st_inspect_pk(path)`** - Validate primary key uniqueness and coverage

### Partitioned and Pruning Data (under development)

- **`st_save_part(x, base, key, ...)`** - Save a single partition with Hive-style directories
- **`st_auto_partition(x, base, partition_cols, ...)`** - Automatically split and save dataset by partition columns
- **`st_load_parts(base, filter = NULL, as = "rbind")`** - Load and combine partitions matching a filter
- **`st_list_parts(base, filter = NULL)`** - List available partitions without loading
- **`st_part_path(base, key)`** - Construct path for a partition given its key
- **`st_prune_versions(path, policy)`** - Remove old versions based on retention policy

### Aliases (Multi-Directory Support)

- **`st_alias_list()`** - List all registered aliases
- **`st_alias_get(name = NULL)`** - Get configuration for an alias

### Builders & Rebuilds ⚠️ *Experimental*

> **Note**: The builder system is under active development. Safe for prototyping, but consider pinning your stamp version in production code until the API stabilizes (expected in v1.0).

- **`st_register_builder(path, builder_fn)`** - Register a function to rebuild an artifact from its parents
- **`st_clear_builders(paths = NULL)`** - Clear registered builders
- **`st_plan_rebuild(targets, ...)`** - Compute rebuild plan (which targets are stale and why)
- **`st_rebuild(plan)`** - Execute a rebuild plan, calling builders and saving results

### Filtering Helpers ⚠️ *Experimental*

> **Note**: Advanced filtering utilities are under development.

- **`st_filter(df, filters = list(), strict = TRUE)`** - Apply named list filters to data frames (used internally for partition queries)

## Learn More

For detailed guides and workflows, see the package vignettes:

- **[Setup and Basics](https://randrescastaneda.github.io/stamp/articles/setup-and-basics.html)** - Getting started with stamp
- **[Hashing and Versions](https://randrescastaneda.github.io/stamp/articles/hashing-and-versions.html)** - Understanding content hashing and version control
- **[Using Aliases](https://randrescastaneda.github.io/stamp/articles/using-alias.html)** - Managing multiple stamp directories
- **[Partitions](https://randrescastaneda.github.io/stamp/articles/partitions.html)** - Working with partitioned datasets
- **[Lineage and Rebuilds](https://randrescastaneda.github.io/stamp/articles/lineage-rebuilds.html)** - Dependency tracking and automated rebuilds
- **[Version Retention](https://randrescastaneda.github.io/stamp/articles/version_retention_prune.html)** - Managing version history with retention policies
- **[Stamp Directory](https://randrescastaneda.github.io/stamp/articles/stamp-directory.html)** - Understanding the `.stamp/` internal structure
