---
title: "Builders, Plans, and Rebuilds"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Builders, Plans, and Rebuilds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE # examples are illustrative; set to TRUE when running interactively
)
```

This short vignette focuses on the core concepts that enable reproducible,
incremental rebuilding with `stamp`: builders, plans, and rebuilds. It complements
`vignettes/stamp.Rmd` by isolating patterns and best practices you should follow
when registering programmatic targets.

```{r setup}
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(".")
} else {
  library(stamp)
}
library(data.table)
set.seed(123)
```

## Key concepts

- Builder: a function registered with `st_register_builder(path, builder_fn)`.
  Signature: `function(path, parents)`. It must return a list with at least:
  - `x`: the object to save
  - `code`: the producing function(s) used to compute `x` (used to compute code_hash)
  - optional `code_label`: a short tag used in sidecars

- Plan: the result of `st_plan_rebuild()` which inspects recorded parents,
  `code_hash` values and artifact presence. It describes actions required to
  bring targets up-to-date.

- Rebuild: executing the plan with `st_rebuild(plan)`. `st_rebuild()` calls the
  registered builders and then `st_save()` for each produced artifact, recording
  new versions and sidecars.

## Minimal example (single-target)

```{r single-target}
# Suppose we have a summary target path
summary_path <- "outputs/welfare_summary.qs2"

# Register a simple builder that loads parents and returns x+code
st_register_builder(summary_path, function(path, parents) {
  # Load welfare parents (illustrative — adapt to your parents layout)
  welfare_paths <- vapply(
    parents,
    function(p) grepl("welfare", p$path),
    logical(1)
  )
  welfare_list <- lapply(parents[welfare_paths], function(p) st_load(p$path))

  # Load macros (shared parents)
  cpi_tbl <- st_load("data/macro/cpi.qs2")
  gdp_tbl <- st_load("data/macro/gdp.qs2")
  pop_tbl <- st_load("data/macro/population.qs2")

  # call a pure worker that accepts pre-loaded inputs
  out <- bar(welfare_list, cpi_tbl, gdp_tbl, pop_tbl)

  list(
    x = out,
    code = list(bar = bar),
    code_label = "aggregate_welfare"
  )
})

# Inspect plan (strict mode is conservative)
plan <- st_plan_rebuild(
  targets = summary_path,
  include_targets = TRUE,
  mode = "strict"
)
plan

# Run only if you're ready
# st_rebuild(plan)
```

## Multi-function `code` metadata

If your target depends on multiple functions (e.g. `bar()` calls helpers
`h1()` and `h2()`), include them explicitly in `code` so `st_save()` computes a
combined hash that changes if any helper changes:

```{r code-multi}
# Good: include all contributing functions
st_save(
  x,
  summary_path,
  pk = c("country", "year"),
  parents = parents,
  code = list(bar = bar, h1 = h1, h2 = h2)
)
```

This ensures `st_is_stale()` and plan computation detect code changes across the
whole chain.

## Modes: strict vs relaxed

- `mode = "strict"` treats missing parents or mismatched code hashes conservatively — prefer this for correctness.
- `mode = "relaxed"` may skip some checks to avoid unnecessary recomputation in noisy environments; use with caution.

## Tips & best practices

- Prefer pure worker functions that accept inputs as arguments. It makes testing
  and registration easier.
- Register builders that use the `parents` argument; avoid global state when
  possible.
- Include helper functions in `code` metadata so any relevant change invalidates
  dependent artifacts.
- Keep builders deterministic and side-effect free (don't write inside builder;
  instead return `x` and let `st_rebuild()` call `st_save()`).

## Try it locally

To try the examples interactively, set `eval = TRUE` in the setup chunk and run
examples inside an interactive R session. Use a short `root_dir` to avoid
Windows PATH length issues.


## Appendix: example plan interpretation

A plan is a list-of-lists; each entry includes `target` (path), a `reason`
(`parent_changed`, `code_changed`, or `missing`) and builder info. Inspect the
plan and filter or reorder actions before calling `st_rebuild()`.

Example (pseudo-inspection):

```{r plan-inspect}
plan <- st_plan_rebuild(targets = summary_path, include_targets = TRUE)
# Show targets that are stale
Filter(function(x) x$reason != "uptodate", plan)
```



