---
title: "Stamp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stamp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# The Problem {.unnumbered}

You have multiple input files produced by different processes. You want
to read, process, and save derived artifacts while capturing.

Let's simulate the data and folder structure:
```{r setup}
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(".")
} else {
  library(stamp)
}
library(data.table)
set.seed(123)

# Create an isolated temporary root so all writes occur outside the package tree.
# Use a short name to avoid long path issues on Windows (PATH_MAX = 260).
root_dir <- fs::path_temp("s") |>
  fs::dir_create()
old_wd <- setwd(root_dir)
on.exit(setwd(old_wd), add = TRUE)

# Initialize stamp under this temp root
st_init(root = root_dir)

# Helper: list only data files with known extensions to avoid locks/sidecars
# Returns relative paths from alias root (suitable for st_load)
data_files <- function(dir) {
  # Convert dir to relative path from root_dir if it's absolute
  rel_dir <- if (fs::is_absolute_path(dir)) {
    as.character(fs::path_rel(dir, start = root_dir))
  } else {
    dir
  }
  
  # Get absolute path for listing
  abs_dir <- fs::path(root_dir, rel_dir)
  if (!fs::dir_exists(abs_dir)) {
    return(character(0))
  }
  
  f <- fs::dir_ls(abs_dir)
  ext <- tolower(fs::path_ext(f))
  allowed <- c("qs2", "rds", "csv", "fst", "json")
  f <- f[ext %in% allowed]
  
  # Return relative paths from root_dir
  as.character(fs::path_rel(f, start = root_dir))
}
```


# The Solution {.unnumbered}


## Harmonizing Economic Indicators with Versioning & Lineage

We walk through a realistic data engineering scenario: several country/year micro datasets and macro indicators must be combined into a standardized output table. We want:

1. Consistent saving, versioning, and timestamp metadata for each input artifact.
2. Explicit primary keys (PK) recorded on disk.
3. A derived summary table saved with full lineage (parents listed).
4. Demonstrating how a single input change propagates using `st_is_stale()`, `st_plan_rebuild()`, and `st_rebuild()`.
5. Adding new data (ARG 2015) and observing minimal recomputation.
6. Producing the same output using partitioned artifacts for partial re-runs.

* Countries/years for welfare (micro) data: COL2010, COL2012, MEX2010, MEX2015, PRY2011, PRY2014.

* Reporting levels:
  - COL and MEX: both `urban` and `rural`.
  - PRY: only `national`.

* Macro data (CPI, GDP, population) exist for COL, MEX, PRY, BRA, ARG (2010–2015). CPI & population have all three reporting levels (`national`, `urban`, `rural`) for every country-year.

## 1. Simulate and Save Input Micro Data

We'll create small synthetic micro datasets with columns:
`country`, `year`, `reporting_level`, `hh_id`, `welfare`, `weight`.

```{r simulate-welfare}
welfare_specs <- data.frame(
  country = c("COL", "COL", "MEX", "MEX", "PRY", "PRY"),
  year = c(2010, 2012, 2010, 2015, 2011, 2014),
  stringsAsFactors = FALSE
)

mk_reporting <- function(cty) {
  if (cty %in% c("COL", "MEX")) c("urban", "rural") else c("national")
}

simulate_welfare <- function(country, year) {
  rl <- mk_reporting(country)
  # 200 households per reporting level
  dt_list <- lapply(rl, function(rlev) {
    n <- 200L
    data.table(
      country = country,
      year = year,
      reporting_level = rlev,
      hh_id = sprintf("%s_%s_%s_%03d", country, year, rlev, seq_len(n)),
      welfare = round(
        rlnorm(n, meanlog = log(1000 + year) - 7, sdlog = 0.5),
        2
      ),
      weight = runif(n, 0.5, 2)
    )
  })
  rbindlist(dt_list)
}

welfare_paths <- list()
for (i in seq_len(nrow(welfare_specs))) {
  row <- welfare_specs[i, ]
  dt <- simulate_welfare(row$country, row$year)
  fn <- sprintf("data/welfare/%s_%d.qs2", row$country, row$year)
  # Primary key columns include household id for uniqueness
  st_save(
    dt,
    fn,
    pk = c("country", "year", "reporting_level", "hh_id"),
    domain = "welfare",
    alias = NULL
  )
  welfare_paths[[length(welfare_paths) + 1]] <- fn
}
unlist(welfare_paths)
```

Each call to `st_save()` assigns a version id (if content differs) and writes sidecar metadata including PK.

Inspect one artifact:

```{r welfare-info}
st_info("data/welfare/COL_2010.qs2", alias = NULL)
```

## 2. Simulate and Save Macro Data (CPI, GDP, Population)

CPI & population are reporting-level granular; GDP is country-year only.

```{r simulate-macro}
years <- 2010:2015
countries_macro <- c("COL", "MEX", "PRY", "BRA", "ARG")
levels_all <- c("national", "urban", "rural")

simulate_cpi <- function() {
  out <- CJ(
    country = countries_macro,
    year = years,
    reporting_level = levels_all
  )
  out[, cpi := round(runif(.N, 80, 140), 2)]
  out
}

simulate_population <- function() {
  out <- CJ(
    country = countries_macro,
    year = years,
    reporting_level = levels_all
  )
  out[, population := round(runif(.N, 5e5, 20e6))]
  out
}

simulate_gdp <- function() {
  out <- CJ(country = countries_macro, year = years)
  out[, gdp := round(runif(.N, 1e9, 5e11))]
  out
}

cpi <- simulate_cpi()
pop <- simulate_population()
gdp <- simulate_gdp()

st_save(
  cpi,
  "data/macro/cpi.qs2",
  pk = c("country", "year", "reporting_level"),
  domain = "macro",
  alias = NULL
)
st_save(
  pop,
  "data/macro/population.qs2",
  pk = c("country", "year", "reporting_level"),
  domain = "macro",
  alias = NULL
)
st_save(
  gdp,
  "data/macro/gdp.qs2",
  pk = c("country", "year"),
  domain = "macro",
  alias = NULL
)
```

Version listing for CPI:

```{r cpi-versions}
print(st_versions("data/macro/cpi.qs2", alias = NULL))
```

## 3. Aggregation Function `foo()` and Output Table

We compute per `(country, year, reporting_level)` welfare statistics and merge with CPI, GDP, population. We demonstrate loading inputs inside the function for self-containment.

```{r define-foo}
foo <- function() {
  # Load all welfare micro artifacts (only recognized data files)
  welfare_files <- data_files("data/welfare")
  if (length(welfare_files) == 0) {
    cat("No welfare files found; returning empty table.\n")
    return(data.table::data.table())
  }
  wf <- rbindlist(lapply(welfare_files, st_load))

  stats <- wf[,
    .(
      welfare_mean = weighted.mean(welfare, weight),
      welfare_median = as.numeric(stats::median(welfare)),
      welfare_sd = stats::sd(welfare),
      welfare_n = .N
    ),
    by = .(country, year, reporting_level)
  ]

  cpi <- st_load("data/macro/cpi.qs2", alias = NULL)
  gdp <- st_load("data/macro/gdp.qs2", alias = NULL)
  pop <- st_load("data/macro/population.qs2", alias = NULL)

  # Merge; CPI & population contain rows not present in welfare; keep welfare rows only
  out <- merge(
    stats,
    cpi,
    by = c("country", "year", "reporting_level"),
    all.x = TRUE
  )
  out <- merge(
    out,
    pop,
    by = c("country", "year", "reporting_level"),
    all.x = TRUE
  )
  out <- merge(out, gdp, by = c("country", "year"), all.x = TRUE)
  setcolorder(
    out,
    c(
      "country",
      "year",
      "reporting_level",
      "welfare_mean",
      "welfare_median",
      "welfare_sd",
      "welfare_n",
      "cpi",
      "gdp",
      "population"
    )
  )
  out
}

# Call foo() to compute summary (empty if no data found)
summary_table <- foo()
if (nrow(summary_table) > 0) {
  head(summary_table)
} else {
  cat("Summary table is empty; ensure welfare data was saved.\n")
  summary_table
}
```

Save the summary with lineage: parents include all welfare micro datasets plus CPI/GDP/population.

```{r save-summary}
# Only save summary if it has data (i.e., welfare data was available)
if (nrow(summary_table) > 0) {
  parent_paths <- c(
    data_files("data/welfare"),
    "data/macro/cpi.qs2",
    "data/macro/gdp.qs2",
    "data/macro/population.qs2"
  )
  parents <- lapply(parent_paths, function(p) {
    list(path = p, version_id = st_latest(p, alias = NULL))
  })

  st_save(
    summary_table,
    "outputs/welfare_summary.qs2",
    pk = c("country", "year", "reporting_level"),
    parents = parents,
    domain = "summary",
    alias = NULL
  )
  st_lineage("outputs/welfare_summary.qs2", depth = 1, alias = NULL)
} else {
  cat("Summary table is empty; skipping save and lineage demo.\n")
}
```

Note: to make provenance explicit, pass the producing function as `code=` so `st_save()` records the function hash. Below we demonstrate both patterns: `foo()` loads its own data, while `bar()` accepts pre-loaded objects and is saved with `code = bar`.

```{r bar-and-save}
# Variant that accepts loaded objects (preferred for testability and clarity)
bar <- function(welfare_list, cpi_tbl, gdp_tbl, pop_tbl) {
  if (length(welfare_list) == 0) {
    cat("No welfare data provided; returning empty table.\n")
    return(data.table::data.table())
  }
  wf <- data.table::rbindlist(welfare_list)
  stats <- wf[,
    .(
      welfare_mean = weighted.mean(welfare, weight),
      welfare_median = as.numeric(stats::median(welfare)),
      welfare_sd = stats::sd(welfare),
      welfare_n = .N
    ),
    by = .(country, year, reporting_level)
  ]

  out <- merge(
    stats,
    cpi_tbl,
    by = c("country", "year", "reporting_level"),
    all.x = TRUE
  )
  out <- merge(
    out,
    pop_tbl,
    by = c("country", "year", "reporting_level"),
    all.x = TRUE
  )
  out <- merge(out, gdp_tbl, by = c("country", "year"), all.x = TRUE)
  setcolorder(
    out,
    c(
      "country",
      "year",
      "reporting_level",
      "welfare_mean",
      "welfare_median",
      "welfare_sd",
      "welfare_n",
      "cpi",
      "gdp",
      "population"
    )
  )
  out
}

# Example: load inputs outside the worker and call bar()
welfare_files <- data_files("data/welfare")
if (length(welfare_files) > 0) {
  welfare_list <- lapply(welfare_files, st_load)
  cpi_tbl <- st_load("data/macro/cpi.qs2", alias = NULL)
  gdp_tbl <- st_load("data/macro/gdp.qs2", alias = NULL)
  pop_tbl <- st_load("data/macro/population.qs2", alias = NULL)

  summary_table2 <- bar(welfare_list, cpi_tbl, gdp_tbl, pop_tbl)
  if (nrow(summary_table2) > 0) {
    parent_paths <- c(
      data_files("data/welfare"),
      "data/macro/cpi.qs2",
      "data/macro/gdp.qs2",
      "data/macro/population.qs2"
    )
    parents <- lapply(parent_paths, function(p) {
      list(path = p, version_id = st_latest(p, alias = NULL))
    })
    st_save(
      summary_table2,
      "outputs/welfare_summary.qs2",
      pk = c("country", "year", "reporting_level"),
      parents = parents,
      code = bar, # addition the function as parent so st_save() can track it
      domain = "summary",
      alias = NULL
    )
    st_lineage("outputs/welfare_summary.qs2", depth = 1, alias = NULL)
  } else {
    cat("Summary table2 is empty; skipping save.\n")
  }
} else {
  cat("No welfare data to process in bar-and-save example.\n")
}
```

Handling multiple function dependencies

If the output depends on more than one function (for example `foo()` calls `f1()` and `f2()`), you can provide a combined `code` metadata to `st_save()` that captures all producing functions. `st_save()` accepts any object for `code`; best practice is to pass a short list of the functions involved, e.g. `code = list(foo = foo, helper1 = f1, helper2 = f2)`. The package hashes `code` and records a `code_hash` in the sidecar so downstream `st_is_stale()` can detect changes in any contributor.

Example pattern:

```{r code-multiple}
# Provide multiple functions to st_save so code_hash changes if any of them change
# st_save(x, path, code = list(foo = foo, helper1 = f1, helper2 = f2))
```

## 4. Modify CPI (COL 2012) → New Version & Staleness

We change CPI for COL 2012 (all reporting levels) to illustrate a new version and stale downstream artifact.

```{r modify-cpi}
cpi2 <- st_load("data/macro/cpi.qs2", alias = NULL)
cpi2[country == "COL" & year == 2012, cpi := cpi * 1.05] # 5% adjustment
st_save(cpi2, "data/macro/cpi.qs2", alias = NULL) # new version recorded
st_versions("data/macro/cpi.qs2", alias = NULL)[1:3]
st_is_stale("outputs/welfare_summary.qs2") # should be TRUE
```

## Builders, plans and rebuilds — concepts and how we use them here

A short glossary and explanation of the concepts used in the examples above:

- **Artifact**: a saved object on disk (one path managed by stamp).

- **Sidecar**: metadata file beside each artifact (PKs, parents, code hash, domain, timestamps).

- **Version**: immutable snapshot; new version id recorded when `st_save()` detects changed content.

- **Parent**: pointer (path + version_id) to an upstream artifact used to produce a downstream artifact.

- **Lineage**: directed graph linking artifacts via parent relationships; inspect with `st_lineage()`.

- **Builder**: function registered with `st_register_builder(path, builder_fn)` that knows how to produce one artifact. Returns at least `x` (object to save) and `code` (function(s) used) so code changes trigger staleness.

- **Plan**: result of `st_plan_rebuild()` describing which targets are stale (parents changed, code changed, or artifact missing) and thus need rebuilding.

- **Rebuild**: executing a plan via `st_rebuild(plan)`; runs builders, saves artifacts, updates sidecars/versions.

General idea
- Register builders for reproducible targets. Builders should be deterministic and return the object to save (do not save directly inside the builder).
- Use `st_plan_rebuild()` to compute what needs to be rebuilt (it inspects recorded parents and code hashes).
- Inspect the plan, then call `st_rebuild()` to execute builders and record new versions.

How this vignette implements those ideas
- We register a builder for `out_summary_path` (see the `st_register_builder()` call below). That builder returns `list(x = foo(), code = foo, code_label = "aggregate_welfare")`.
- `st_plan_rebuild()` compares current parent version ids and the code hash of `foo` against what the sidecar recorded; if any parent or code changed the plan will mark the target as stale.
- `st_rebuild(plan)` calls the registered builder, then `st_save()` persistently records the new artifact and sidecar (including the new `code_hash` and parent pointers).
- This flow gives you a lightweight, auditable, and incremental pipeline: change an upstream artifact or a function and the plan will tell you what to rebuild.

Practical tips and a small example
- Builders should prefer to use the `parents` argument passed in (this avoids global state and makes builders easier to test).
- If your target depends on multiple functions, pass them all in `code` so `st_save()` records a combined code hash: `code = list(foo = foo, helper = f1)`.

Example builder that loads parents explicitly and returns multiple functions in `code`:

```{r}
# Example: register a builder that uses the provided `parents` list and avoids global loads
st_register_builder("outputs/welfare_summary.qs2", function(path, parents) {
  # parents is a list of lists: each element has $path and $version_id
  # Filter parent paths to identify welfare partitions vs macros, then load them
  welfare_parent_paths <- vapply(
    parents,
    function(p) grepl("welfare", p$path),
    logical(1)
  )
  welfare_list <- lapply(parents[welfare_parent_paths], function(p) {
    st_load(p$path, alias = NULL)
  })

  cpi_tbl <- st_load("data/macro/cpi.qs2", alias = NULL)
  gdp_tbl <- st_load("data/macro/gdp.qs2", alias = NULL)
  pop_tbl <- st_load("data/macro/population.qs2", alias = NULL)

  # Produce the output using a helper function (bar) that accepts pre-loaded inputs
  out <- bar(welfare_list, cpi_tbl, gdp_tbl, pop_tbl)

  # Return the produced object plus code metadata (capture both producer and helpers)
  list(
    x = out,
    code = list(bar = bar), # include helpers as needed
    code_label = "aggregate_welfare_partitioned"
  )
})
```

Inspect the plan before running it:
```{r}
plan <- st_plan_rebuild(
  targets = "outputs/welfare_summary.qs2",
  include_targets = TRUE,
  mode = "strict"
)
print(plan)
```

Run the plan to execute the builders and record new versions:
```{r}
st_rebuild(plan)
```

This example shows the recommended pattern: load from `parents`, call a pure worker (like `bar()`), and return `x` and `code`. That ensures `st_rebuild()` can uniformly save artifacts, record lineage, and compute `code_hash` changes for future incremental rebuilds.

Plan & rebuild using a registered builder for the summary artifact.

```{r rebuild-summary}
st_register_builder("outputs/welfare_summary.qs2", function(path, parents) {
  list(x = foo(), code = foo, code_label = "aggregate_welfare")
})

plan <- st_plan_rebuild(
  targets = "outputs/welfare_summary.qs2",
  include_targets = TRUE,
  mode = "strict"
)
plan
st_rebuild(plan)
st_lineage("outputs/welfare_summary.qs2", depth = 1, alias = NULL)
```

## 5. Add New Data (ARG 2015 Welfare)

We introduce a new welfare dataset (ARG 2015) which appears in macro tables already. Only the summary rows for ARG 2015 will be newly added.

```{r add-arg}
dt_arg_2015 <- simulate_welfare("ARG", 2015)
st_save(
  dt_arg_2015,
  "data/welfare/ARG_2015.qs2",
  pk = c("country", "year", "reporting_level", "hh_id"),
  domain = "welfare",
  alias = NULL
)

# Only recompute and save summary if we have welfare data
if (length(data_files("data/welfare")) > 0 && nrow(summary_table) > 0) {
  parents2 <- lapply(
    c(
      data_files("data/welfare"),
      "data/macro/cpi.qs2",
      "data/macro/gdp.qs2",
      "data/macro/population.qs2"
    ),
    function(p) list(path = p, version_id = st_latest(p, alias = NULL))
  )
  summary_table_new <- foo()
  if (nrow(summary_table_new) > 0) {
    st_save(
      summary_table_new,
      "outputs/welfare_summary.qs2",
      pk = c("country", "year", "reporting_level"),
      parents = parents2,
      domain = "summary",
      alias = NULL
    )
    head(st_load("outputs/welfare_summary.qs2", alias = NULL))
  } else {
    cat("Summary table after adding ARG 2015 is empty; skipping save.\n")
  }
} else {
  cat("Insufficient data to recompute summary.\n")
}
```

## 6. Partitioned Workflow for Partial Re-execution

Instead of storing one large welfare micro file per country-year, we can store partitions per `(country, year, reporting_level)` using `st_save_part()` and then recompute only the affected partitions when an input changes.

```{r partition-save}
# Write partitioned welfare (one partition per country/year/reporting_level)
for (i in seq_len(nrow(welfare_specs))) {
  row <- welfare_specs[i, ]
  dt <- simulate_welfare(row$country, row$year) # new simulated (independent from earlier saves)
  # split by reporting_level
  for (rl in unique(dt$reporting_level)) {
    part_dt <- dt[reporting_level == rl]
    st_save_part(
      part_dt,
      base = "data/welfare_parts",
      key = list(country = row$country, year = row$year, reporting_level = rl),
      code_label = "welfare_partition",
      pk = c("country", "year", "reporting_level", "hh_id"),
      alias = NULL
    )
  }
}
st_list_parts("data/welfare_parts")[1:6, ]
```

Define a partitioned builder writing one output partition per `(country, year, reporting_level)` summary row. Each output partition depends on: the corresponding welfare partition + CPI + GDP + population (all three macro artifacts are shared parents).

```{r partition-builder}

build_partition_summary <- function(path, parents) {
  welfare_parts_dir <- fs::path(root_dir, "data", "welfare_parts")
  summary_parts_dir <- fs::path(root_dir, "outputs", "summary_parts")
  
  # path is the final destination for this partition artifact
  # Extract key by parsing path (we stored key in directory names)
  rel <- fs::path_rel(path, start = summary_parts_dir)
  segs <- strsplit(dirname(rel), .Platform$file.sep)[[1]]
  segs <- segs[nzchar(segs)]
  # Extract only segments that follow the key=value pattern and ignore others
  kv_pairs <- lapply(segs, function(s) {
    m <- regmatches(s, regexec("^([^=]+)=(.*)$", s))[[1]]
    if (length(m) == 3L) list(k = m[2], v = m[3]) else NULL
  })
  kv_pairs <- Filter(Negate(is.null), kv_pairs)
  if (!length(kv_pairs)) {
    key <- list()
  } else {
    key <- setNames(
      lapply(kv_pairs, function(x) x$v),
      vapply(kv_pairs, function(x) x$k, character(1))
    )
  }

  # Load welfare partition matching key
  w_path <- st_part_path("data/welfare_parts", key, format = NULL, alias = NULL)
  w <- st_load(w_path, alias = NULL)

  stats <- w[,
    .(
      welfare_mean = weighted.mean(welfare, weight),
      welfare_median = as.numeric(median(welfare)),
      welfare_sd = sd(welfare),
      welfare_n = .N
    ),
    by = .(country, year, reporting_level)
  ]

  # Load macro (shared) and subset to key
  cpi <- st_load("data/macro/cpi.qs2", alias = NULL)[
    country == key$country &
      year == as.integer(key$year) &
      reporting_level == key$reporting_level
  ]
  pop <- st_load("data/macro/population.qs2", alias = NULL)[
    country == key$country &
      year == as.integer(key$year) &
      reporting_level == key$reporting_level
  ]
  gdp <- st_load("data/macro/gdp.qs2", alias = NULL)[
    country == key$country & year == as.integer(key$year)
  ]

  out <- merge(
    stats,
    cpi,
    by = c("country", "year", "reporting_level"),
    all.x = TRUE
  )
  out <- merge(
    out,
    pop,
    by = c("country", "year", "reporting_level"),
    all.x = TRUE
  )
  out <- merge(out, gdp, by = c("country", "year"), all.x = TRUE)
  list(
    x = out,
    code = build_partition_summary,
    code_label = "partition_summary"
  )
}
```

```{r partition-summary-build}
# Populate outputs/summary_parts by running the builder for each welfare partition key
summary_parts_dir <- fs::path(root_dir, "outputs", "summary_parts")
for (i in seq_len(nrow(partition_keys))) {
  key_row <- partition_keys[i, ]
  # Reconstruct key list from the partition_keys table (drop non-key columns)
  key_cols <- setdiff(names(key_row), "path")
  key <- as.list(key_row[, key_cols, drop = FALSE])
  key <- Filter(Negate(is.na), key)

  # Construct the output partition path
  out_path <- st_part_path(
    "outputs/summary_parts",
    key,
    format = "qs2"
  )

  # Build parents list: welfare partition + the three macro files
  welfare_path <- st_part_path("data/welfare_parts", key, format = NULL)
  parents_i <- list(
    list(path = welfare_path,              version_id = st_latest(welfare_path,              alias = NULL)),
    list(path = "data/macro/cpi.qs2",        version_id = st_latest("data/macro/cpi.qs2",        alias = NULL)),
    list(path = "data/macro/gdp.qs2",        version_id = st_latest("data/macro/gdp.qs2",        alias = NULL)),
    list(path = "data/macro/population.qs2", version_id = st_latest("data/macro/population.qs2", alias = NULL))
  )

  result <- build_partition_summary(out_path, parents_i)
  st_save(
    result$x,
    out_path,
    parents = parents_i,
    pk = c("country", "year", "reporting_level"),
    code = result$code,
    code_label = result$code_label,
    alias = NULL
  )
}
```

```{r partition-summary-list}
# List a few partitioned summary partitions
st_list_parts("outputs/summary_parts")[1:6, ]
```

### Partial Re-execution After CPI Change

If CPI changes only for COL 2012, we want to rebuild only partitions for COL 2012 across reporting levels.

```{r partition-update-cpi}
# Modify CPI again (COL 2012) to trigger staleness
cpi3 <- st_load("data/macro/cpi.qs2", alias = NULL)
cpi3[country == "COL" & year == 2012, cpi := cpi * 1.02]
st_save(cpi3, "data/macro/cpi.qs2", alias = NULL)

# Detect which partition outputs are stale (simple check: those whose parent CPI version differs)
summary_parts <- st_list_parts("outputs/summary_parts")
stale_idx <- vapply(summary_parts$path, st_is_stale, logical(1))
summary_parts[stale_idx, ]
```

Rebuild only stale partitions (manually filtered):

```{r partition-rebuild-stale}
stale_paths <- summary_parts$path[stale_idx]
summary_parts_dir <- fs::path(root_dir, "outputs", "summary_parts")
for (p in stale_paths) {
  # reconstruct key from path to identify its welfare partition parent
  rel <- fs::path_rel(p, start = summary_parts_dir)
  segs <- strsplit(dirname(rel), .Platform$file.sep)[[1]]
  kv <- lapply(segs, function(s) strsplit(s, "=")[[1]])
  key <- setNames(lapply(kv, function(x) x[2]), lapply(kv, function(x) x[1]))
  parents <- list(
    list(
      path = st_part_path("data/welfare_parts", key, alias = NULL),
      version_id = st_latest(st_part_path("data/welfare_parts", key, alias = NULL), alias = NULL)
    ),
    list(
      path = "data/macro/cpi.qs2",
      version_id = st_latest("data/macro/cpi.qs2", alias = NULL)
    ),
    list(
      path = "data/macro/gdp.qs2",
      version_id = st_latest("data/macro/gdp.qs2", alias = NULL)
    ),
    list(
      path = "data/macro/population.qs2",
      version_id = st_latest("data/macro/population.qs2", alias = NULL)
    )
  )
  b <- build_partition_summary(p, parents)
  st_save(
    b$x,
    p,
    parents = parents,
    pk = c("country", "year", "reporting_level"),
    code = b$code,
    code_label = b$code_label,
    alias = NULL
  )
}
```

Only COL 2012 partitions were recomputed, avoiding unnecessary work for other countries/years.

## 7. Summary

We demonstrated:

* Saving multiple artifacts with PK metadata and automatic versioning.
* Building a derived summary with explicit lineage and rebuilding after an upstream change.
* Adding new data (ARG 2015) and incorporating it with minimal friction.
* A partitioned strategy enabling targeted re-execution for changed inputs.

Explore lineage further:

```{r final-lineage}
st_lineage("outputs/welfare_summary.qs2", depth = 2, alias = NULL)[1:10, ]
```

<warning>
**Warning: The code below will permanently delete the temporary vignette root directory and all its contents.**
</warning>

```{r cleanup, eval= interactive()}
# Remove the temporary root used for vignette examples. Only run in a local
# interactive session when you are sure you no longer need the temporary files.
if (exists("root_dir") && fs::dir_exists(root_dir)) {
  # fs::dir_delete() permanently removes the directory tree.
  fs::dir_delete(root_dir)
}
```
