---
title: "The .stamp Directory: Structure and Internals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The .stamp Directory: Structure and Internals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(".")
} else {
  library(stamp)
}
```

## Overview

The `.stamp/` directory is the **persistent storage backend** for the `stamp` package's versioning system. This vignette explains:

1. How and when `.stamp/` is created
2. Its internal directory structure
3. How versioning and metadata work under the hood
4. Troubleshooting common issues

**Audience:** This vignette serves both **users** (understanding what `.stamp` does) and **developers** (understanding internal implementation details).

> Architecture note
>
> Recent changes introduced alias-aware paths and a dedicated storage directory `.st_data/` under the project root. In this model:
> - `.stamp/` continues to store the catalog, locks, and immutable version snapshots.
> - `.st_data/` can hold artifact files when using storage-managed paths, while your code still refers to logical paths.
> - Paths passed to APIs must resolve under the current alias root (the directory you called `st_init()` on). Prefer saving within that root (e.g., `fs::path(demo_dir, ...)`).
>
> The examples below use in-project paths to keep the focus on `.stamp/` internals.

---

## 1. Creation and Initialization

### 1.1 Creating `.stamp/` with `st_init()`

The `.stamp/` directory is created when you call `st_init()`:

```{r init-demo}
# Create a temporary project directory for demonstration
demo_dir <- fs::path_temp("stamp-demo")
fs::dir_create(demo_dir)

# Initialize stamp
st_init(demo_dir)

# Inspect what was created
fs::dir_tree(fs::path(demo_dir, ".stamp"), recurse = TRUE, all = TRUE)
```

**What happens during initialization:**

1. **Creates directory structure** (if it doesn't exist):
   - `.stamp/` - Root state directory
   - `.stamp/temp/` - Temporary files during atomic writes
   - `.stamp/logs/` - Future use for logging (currently unused)

2. **Records project root** in package state (in-memory reference)

3. **Does NOT create or initialize** the catalog yet - that happens on first save

### 1.2 Re-running `st_init()`: Safe and Non-Destructive

**Important:** Running `st_init()` multiple times is **safe** and will **NOT** delete or overwrite existing version history.

```{r reinit-demo}
# Save an artifact to create version history
test_data <- data.frame(x = 1:5, y = letters[1:5])
test_path <- fs::path(demo_dir, "data", "test.qs2")
st_save(test_data, test_path)

# Check versions exist
versions_before <- st_versions(test_path)
nrow(versions_before)

# Re-initialize (this is safe!)
st_init(demo_dir)

# Version history is preserved
versions_after <- st_versions(test_path)
identical(versions_before, versions_after)
```

**Why is this safe?**

- `st_init()` only creates directories that don't exist
- The catalog file (`catalog.qs2`) is read if present, never deleted
- Version snapshots in `versions/` are never touched by initialization

---

## 2. Directory Structure Explained

### 2.1 High-Level Layout

```
<project-root>/
├── .stamp/                      # State directory (catalog, locks, temp)
│   ├── catalog.qs2              # Central version registry
│   ├── catalog.lock             # Lock file for concurrent access
│   ├── temp/                    # Temporary files during atomic writes
│   └── logs/                    # Reserved for future logging
│
├── results/                     # Your project structure (example)
│   └── model.rds/               # Artifact directory
│       ├── model.rds            # The actual artifact file
│       ├── stmeta/              # Sidecar metadata directory
│       │   ├── sidecar.json     # JSON metadata
│       │   └── sidecar.qs2      # (optional) Binary metadata
│       └── versions/            # Version history (per-artifact)
│           ├── 20250108T121500Z-abc12345/   # Version snapshot
│           │   ├── artifact                  # Snapshot of the file
│           │   ├── sidecar.json              # Metadata at save time
│           │   ├── sidecar.qs2               # (optional) Binary metadata
│           │   └── parents.json              # Lineage information
│           └── 20250108T143000Z-def67890/   # Another version
│               └── ...
│
└── data.qs2/                    # Bare filename (no subdirectory path)
    ├── data.qs2                 # The actual artifact file
    ├── stmeta/                  # Sidecar metadata
    └── versions/                # Version history for this artifact
```

Let's explore each component:

### 2.2 State Directory: `.stamp/`

The `.stamp/` directory contains **only state** - no artifact data:

- **`catalog.qs2`** - Central registry of all artifacts and versions
- **`catalog.lock`** - Concurrency control for catalog updates
- **`temp/`** - Temporary files during atomic writes
- **`logs/`** - Reserved for future use

**Key change (v0.0.9):** Version snapshots are **no longer** under `.stamp/versions/`. They now live alongside each artifact in per-artifact `versions/` directories.

### 2.3 The Catalog: `catalog.qs2`

The catalog is a **central registry** tracking all artifacts and their versions. It's a QS2-serialized list containing two `data.table` objects:

```{r catalog-structure, eval=FALSE}
catalog <- list(
  artifacts = data.table(
    artifact_id,          # Stable hash of normalized path
    path,                 # Current canonical path
    format,               # File format (rds, qs2, csv, etc.)
    latest_version_id,    # Most recent version identifier
    n_versions            # Total number of saved versions
  ),
  versions = data.table(
    version_id,           # Unique version identifier
    artifact_id,          # Links to artifacts table
    content_hash,         # Hash of file contents
    code_hash,            # Hash of generating code (if tracked)
    size_bytes,           # File size in bytes
    created_at,           # ISO8601 UTC timestamp
    sidecar_format        # "json", "qs2", "both", or "none"
  )
)
```

**Key functions that read the catalog:**

```{r catalog-functions}
# List all versions of an artifact
versions <- st_versions(test_path)
str(versions)

# Get latest version ID
latest_id <- st_latest(test_path)
latest_id

# Get comprehensive info (catalog + sidecar + snapshot location)
info <- st_info(test_path)
str(info, max.level = 1)
```

### 2.4 Version Snapshots: Per-Artifact `versions/`

Each time you save an artifact (and versioning is enabled), a new **immutable snapshot** is created in a `versions/` directory **next to the artifact itself**:

```{r version-snapshot-demo}
# Save the same artifact multiple times with changes
st_opts(versioning = "timestamp")  # ensure snapshots are created
v1 <- data.frame(x = 1:3)
st_save(v1, test_path, code_label = "initial")
Sys.sleep(1.1)

v2 <- data.frame(x = 1:5)
st_save(v2, test_path, code_label = "added rows")
Sys.sleep(1.1)

v3 <- data.frame(x = 1:5, y = 10:14)
st_save(v3, test_path, code_label = "added column")

# Version history is stored NEXT TO the artifact, not in .stamp/
# Extract the artifact directory (e.g., demo_dir/data/test.qs2/)
artifact_dir <- fs::path_dir(test_path)
versions_dir <- fs::path(artifact_dir, "versions")

if (fs::dir_exists(versions_dir)) {
  cat("Version snapshots location:", versions_dir, "\n")
  fs::dir_tree(versions_dir, recurse = 1)
} else {
  cat("No versions directory found (versioning may be off).\n")
}

# reset to default behavior for the rest of the vignette
st_opts(versioning = "content")
```

**What's inside a version snapshot directory?**

Each version snapshot (e.g., `<artifact-dir>/versions/20250108T121500Z-abc12345/`) contains:

1. **`artifact`** - A complete copy of the file at that point in time
2. **`sidecar.json` / `sidecar.qs2`** - Metadata including:
   - Content hash
   - Code hash (if tracked)
   - File size
   - Timestamp
   - Custom metadata tags
   - Parent references (quick view)
3. **`parents.json`** - Immutable provenance chain showing which artifacts this version depends on (only present if parents were specified)

```{r inspect-snapshot}
# Get the latest version directory path
latest_info <- st_info(test_path)
latest_vdir <- latest_info$snapshot_dir

if (!is.na(latest_vdir) && fs::dir_exists(latest_vdir)) {
  cat("Latest snapshot directory:", latest_vdir, "\n\n")
  
  # List contents - note: parents.json only exists if parents were specified
  fs::dir_ls(latest_vdir)
  
  # Read the sidecar from the snapshot
  sidecar_path <- fs::path(latest_vdir, "sidecar.json")
  if (fs::file_exists(sidecar_path)) {
    sidecar <- jsonlite::read_json(sidecar_path)
    str(sidecar[c("path", "format", "created_at", "size_bytes", "code_label")])
  }
} else {
  cat("No snapshot directory recorded for test_path; ensure versioning created snapshots.\n")
}
```

**Example of specifying parents during save:**

Notice that the `parents.json` file is not present in the example above. This is because it is only created when parents are specified.

```{r parents-demo}
# Ensure snapshots are recorded for this demo
st_opts(versioning = "timestamp")

# First, create an upstream artifact
upstream_path <- fs::path(demo_dir, "data", "upstream.qs2")
upstream_data <- data.frame(id = 1:10, value = rnorm(10))
st_save(upstream_data, upstream_path, code_label = "upstream data")
upstream_version <- st_latest(upstream_path)

# Now create a derived artifact with parent reference
derived_path <- fs::path(demo_dir, "data", "derived.qs2")
derived_data <- data.frame(id = 1:10, transformed = upstream_data$value * 2)
st_save(
  derived_data, 
  derived_path,
  parents = list(list(path = upstream_path, version_id = upstream_version)),
  code_label = "derived from upstream"
)

# Now check the derived artifact's snapshot - parents.json will be present
derived_info <- st_info(derived_path)
derived_vdir <- derived_info$snapshot_dir
if (!is.na(derived_vdir) && fs::dir_exists(derived_vdir)) {
  fs::dir_ls(derived_vdir)
  
  # Read parents.json
  parents_file <- fs::path(derived_vdir, "parents.json")
  if (fs::file_exists(parents_file)) {
    parents <- jsonlite::read_json(parents_file)
    str(parents)
  } else {
    cat("parents.json not found; ensure parents were recorded.\n")
  }
} else {
  cat("No snapshot directory recorded for derived_path; ensure versioning created snapshots.\n")
}

# Reset to default versioning for the remainder
st_opts(versioning = "content")
```

### 2.5 Artifact Storage: Direct-Path Model

**Key concept (v0.0.9):** Artifacts are stored **directly** at the path you specify, wrapped in a parent directory:

```
st_save(data, "results/model.rds")  →  <root>/results/model.rds/model.rds
st_save(data, "data.qs2")           →  <root>/data.qs2/data.qs2
```

This creates a predictable structure:

```
<root>/
├── results/
│   └── model.rds/              # Artifact directory
│       ├── model.rds           # The actual file
│       ├── stmeta/             # Sidecar metadata
│       └── versions/           # Version history for THIS artifact
│           ├── 20250108T121500Z-abc12345/
│           └── 20250108T143000Z-def67890/
│
└── data.qs2/                   # Bare filename
    ├── data.qs2                # The actual file
    ├── stmeta/                 # Sidecar metadata
    └── versions/               # Version history for THIS artifact
        └── 20250108T120000Z-xyz98765/
```

#### Why This Structure?

1. **Transparency**: Artifacts live where you expect them - at the path you specified
2. **Co-location**: Version history, metadata, and the artifact itself are grouped together
3. **Simplicity**: No need to search `.stamp/versions/` - everything is in one place
4. **Distributed**: Each artifact manages its own version history independently

#### Real-World Example

```{r artifact-organization-demo, eval=FALSE}
# Save to a subdirectory path
project_file <- fs::path(demo_dir, "outputs", "results.qs2")
st_save(data.frame(x = 1:5), project_file)

# Creates:
#   <demo_dir>/outputs/results.qs2/results.qs2  (the file)
#   <demo_dir>/outputs/results.qs2/stmeta/      (metadata)
#   <demo_dir>/outputs/results.qs2/versions/    (history)

# Save a bare filename
bare_file <- fs::path(demo_dir, "summary.qs2")
st_save(data.frame(y = 6:10), bare_file)

# Creates:
#   <demo_dir>/summary.qs2/summary.qs2  (the file)
#   <demo_dir>/summary.qs2/stmeta/      (metadata)
#   <demo_dir>/summary.qs2/versions/    (history)
```

#### Loading Artifacts

When you load, just use the original path you saved with:

```{r load-example, eval=FALSE}
# Load using the same path you saved with
data <- st_load("outputs/results.qs2")  # stamp finds outputs/results.qs2/results.qs2
data <- st_load("summary.qs2")          # stamp finds summary.qs2/summary.qs2
```

Stamp automatically resolves the path to find the actual file in its directory wrapper.

---

## 3. How Versioning Works

### 3.1 The Version Creation Process

When you call `st_save()`, here's what happens internally:

```
┌─────────────────────────────────────────────────────────────────┐
│                       st_save(data, path)                       │
└─────────────────────────┬───────────────────────────────────────┘
                          │
                          ▼
         ┌────────────────────────────────────────┐
         │  1. Decide if save is needed           │
         │     (st_should_save)                   │
         ├────────────────────────────────────────┤
         │  • Check if file exists                │
         │  • Compare content hash                │
         │  • Compare code hash                   │
         │  • Check versioning policy             │
         └────────────────┬───────────────────────┘
                          │
                          ▼
         ┌────────────────────────────────────────┐
         │  2. Write artifact atomically          │
         ├────────────────────────────────────────┤
         │  • Write to temp file                  │
         │  • Move to destination (atomic)        │
         └────────────────┬───────────────────────┘
                          │
                          ▼
         ┌────────────────────────────────────────┐
         │  3. Write sidecar metadata             │
         ├────────────────────────────────────────┤
         │  • Compute hashes                      │
         │  • Record timestamp                    │
         │  • Store in stmeta/ directory          │
         └────────────────┬───────────────────────┘
                          │
                          ▼
         ┌────────────────────────────────────────┐
         │  4. Update catalog                     │
         ├────────────────────────────────────────┤
         │  • Add version row to catalog          │
         │  • Update artifact's latest_version_id │
         │  • Increment n_versions counter        │
         └────────────────┬───────────────────────┘
                          │
                          ▼
         ┌────────────────────────────────────────┐
         │  5. Create version snapshot            │
         ├────────────────────────────────────────┤
         │  • Copy artifact to versions/          │
         │  • Copy sidecars                       │
         │  • Write parents.json                  │
         └────────────────┬───────────────────────┘
                          │
                          ▼
         ┌────────────────────────────────────────┐
         │  6. Apply retention policy             │
         │     (if configured)                    │
         ├────────────────────────────────────────┤
         │  • Prune old versions based on policy  │
         └────────────────────────────────────────┘
```

Each step is designed to be **atomic and crash-safe**, ensuring that partial writes never corrupt your version history.

### 3.2 Version Identifiers

Version IDs are deterministic hashes combining:

- Timestamp (microsecond precision)
- Content hash
- Code hash (if available)
- Artifact ID

Example: `20250108T121507123456Z-abc12345`

This ensures:

- **Chronological ordering** - Timestamps sort naturally
- **Uniqueness** - Hash suffix prevents collisions
- **Traceability** - Hash links to specific content state

### 3.3 Versioning Modes

Control versioning behavior with `st_opts()`:

```{r versioning-modes}
# Show current versioning mode
st_opts("versioning", .get = TRUE)

# Available modes:
# - "content" (default): Save version only when content or code changes
# - "timestamp": Save version on every st_save() call
# - "off": Disable versioning (only update current file + sidecar)

# Example: Force version on every save
st_opts(versioning = "timestamp")
v_same <- data.frame(x = 1:3)
st_save(v_same, test_path, code_label = "first")
Sys.sleep(0.2)
st_save(v_same, test_path, code_label = "second identical")  # Still creates version!

# Check: two versions with identical content
recent_versions <- st_versions(test_path)
tail(recent_versions[, .(version_id, created_at, content_hash)], 2)

# Reset to default
st_opts(versioning = "content")
```

---

## 4. Developer Details

### 4.1 Key Internal Functions

These functions power the `.stamp/` infrastructure (from `R/version_store.R`):

**Path and ID Management:**
```{r internal-funcs, eval=FALSE}
.st_norm_path(path)           # Normalize path to absolute canonical form
.st_artifact_id(path)         # Compute stable hash identifier from path
.st_root_dir()                # Get project root from st_init()
.st_state_dir_abs()           # Get absolute .stamp/ path
```

**Catalog Operations:**
```{r catalog-ops, eval=FALSE}
.st_catalog_path()            # Path to catalog.qs2
.st_catalog_read()            # Read catalog (or create empty if missing)
.st_catalog_write(cat)        # Atomic catalog write with locking
.st_catalog_record_version()  # Add new version row to catalog
```

**Version Management:**
```{r version-ops, eval=FALSE}
.st_version_dir(rel_path, vid, alias)    # Compute specific version snapshot path
.st_version_commit_files()    # Copy artifact + sidecars to snapshot
.st_version_read_parents()    # Read parents.json from snapshot
.st_version_write_parents()   # Write parents.json to snapshot
# Note: .st_versions_root() is deprecated (centralized storage removed in v0.0.9)
```

### 4.2 Concurrency and Safety

#### File-Based Locking Explained

**The Problem:** Imagine two R sessions running simultaneously, both trying to save artifacts:

```
Session A: Reads catalog → Modifies → Writes back
Session B: Reads catalog → Modifies → Writes back
```

Without coordination, Session B might overwrite Session A's changes, **losing version records**.

**The Solution:** `stamp` uses **file-based locking** to ensure only one process modifies the catalog at a time:

```
Session A: Acquires lock → Reads → Modifies → Writes → Releases lock
Session B: Waits for lock → Acquires lock → Reads → Modifies → Writes → Releases lock
```

This is implemented via a lock file (`.stamp/catalog.lock`):

```{r locking-demo, eval=FALSE}
# Internal locking mechanism (simplified)
.st_with_lock(path, {
  cat <- .st_catalog_read()     # Read catalog safely
  # ... modify catalog ...       # Make changes
  .st_catalog_write(cat)         # Write back safely
})

# Lock file: .stamp/catalog.lock
# - Created automatically during catalog updates
# - Uses filelock package if available (recommended)
# - 5-second timeout prevents permanent deadlocks
# - Automatically cleaned up after operation completes
```

**Real-world scenarios where locking matters:**

1. **Parallel processing:** Running `future::plan(multisession)` with multiple workers saving artifacts
2. **Shared filesystems:** Multiple analysts on a server saving to the same project
3. **Background jobs:** RStudio background jobs running `st_save()` while you work interactively

**What happens without the filelock package?**

- `stamp` falls back to **advisory locking** (no enforcement, relies on cooperation)
- Risk of corruption increases in high-concurrency scenarios
- Install `filelock` for production use: `install.packages("filelock")`

#### Atomic Operations

Beyond locking, `stamp` uses **atomic operations** to prevent partial writes that could corrupt your data.

**What "atomic" means:** An operation that either completes entirely or fails entirely, with no in-between state visible to other processes.

**Why this matters:** Consider what could go wrong without atomicity:

```
# Bad scenario (non-atomic):
1. Start writing new data to file
2. ❌ CRASH (power outage, R session killed, etc.)
3. File now contains partial/corrupt data
4. Version history references a broken file
```

**How stamp ensures atomicity:**

- **File writes:** Always write to temp file → move to final location (filesystem-level atomic operation)
- **Catalog updates:** Read-modify-write under lock ensures serialized updates
- **Version snapshots:** Immutable once created (copy operations, never modified)

---

## 5. Inspecting `.stamp/` Programmatically

### 5.1 User-Level Inspection

```{r user-inspection}
# Ensure at least two versions exist for demonstration
# Use explicit alias = NULL to auto-detect current alias
versions <- st_versions(test_path, alias = NULL)
if (nrow(versions) < 2) {
  st_opts(versioning = "timestamp")
  tmp <- data.frame(x = seq_len(3L))
  st_save(tmp, test_path, code_label = "autocreate v1 (user-inspection)")
  Sys.sleep(1.1)
  tmp2 <- transform(tmp, y = x * 10L)
  st_save(tmp2, test_path, code_label = "autocreate v2 (user-inspection)")
  versions <- st_versions(test_path, alias = NULL)
  st_opts(versioning = "content")
}

# List all versions of an artifact (compact view)
versions[, .(version_id, created_at, size_bytes)]

# Get comprehensive info
info <- st_info(test_path)
info$catalog      # Latest version and count
info$snapshot_dir # Path to latest snapshot

# Load a specific historical version (previous), safely
if (nrow(versions) > 1) {
  old_version_try <- try(st_load(test_path, version = -1, alias = NULL), silent = TRUE)
  if (!inherits(old_version_try, "try-error")) {
    str(old_version_try)
  } else {
    cat("Previous version not available; skipping load.\n")
  }
}
```

### 5.2 Direct Catalog Access (Advanced)

```{r direct-catalog, eval=FALSE}
# NOT recommended for users, but useful for debugging
catalog_path <- fs::path(demo_dir, ".stamp", "catalog.qs2")
if (fs::file_exists(catalog_path)) {
  cat <- qs2::qs_read(catalog_path)
  
  # View all artifacts
  print(cat$artifacts)
  
  # View all versions
  print(cat$versions)
}
```

### 5.3 Exploring Snapshots

```{r explore-snapshots}
# Get all version directories for an artifact
# Version history is now stored NEXT TO the artifact, not in .stamp/
artifact_dir <- fs::path_dir(test_path)  # e.g., demo_dir/data/test.qs2/
versions_dir <- fs::path(artifact_dir, "versions")

if (fs::dir_exists(versions_dir)) {
  cat("Versions directory:", versions_dir, "\n\n")
  
  # List all version snapshots
  snapshot_dirs <- fs::dir_ls(versions_dir, type = "directory")
  cat("Found", length(snapshot_dirs), "version snapshots\n\n")
  
  # Inspect contents of latest snapshot
  if (length(snapshot_dirs) > 0) {
    latest <- snapshot_dirs[length(snapshot_dirs)]
    cat("Latest snapshot:", latest, "\n")
    fs::dir_tree(latest)
    
    # Read parents.json if present
    parents_file <- fs::path(latest, "parents.json")
    if (fs::file_exists(parents_file)) {
      parents <- jsonlite::read_json(parents_file)
      str(parents)
    }
  }
} else {
  cat("No versions directory found for this artifact\n")
}
```

---

## 6. Troubleshooting

### 6.1 Missing or Corrupt `.stamp/` Directory

**Symptom:** Functions like `st_versions()` return empty results or error.

**Diagnosis:**
```{r diagnose-missing, eval=FALSE}
# Check if .stamp exists
stamp_dir <- fs::path(demo_dir, ".stamp")
fs::dir_exists(stamp_dir)

# Check if catalog exists
catalog_path <- fs::path(stamp_dir, "catalog.qs2")
fs::file_exists(catalog_path)
```

**Solution:**
```{r fix-missing, eval=FALSE}
# Re-initialize (safe, won't delete existing data)
st_init(demo_dir)

# If catalog is corrupt, it will be recreated empty on first st_save()
# Note: This means version history is lost - restore from backup if available
```

### 6.2 Catalog Schema Mismatch

**Symptom:** Error: "Catalog schema mismatch in versions table."

**Cause:** Package upgrade changed catalog structure, or manual corruption.

**Solution:**
```{r fix-schema, eval=FALSE}
# Back up existing catalog
backup_path <- fs::path(stamp_dir, "catalog_backup.qs2")
fs::file_copy(catalog_path, backup_path, overwrite = TRUE)

# Option 1: Delete catalog and rebuild (LOSES VERSION HISTORY)
fs::file_delete(catalog_path)
# Next st_save() will create fresh catalog

# Option 2: Manual migration (advanced - contact package maintainer)
```

### 6.3 Disk Space Issues

**Symptom:** Version history grows very large across many artifacts.

**Diagnosis:**
```{r disk-usage, eval=FALSE}
# Check total .stamp/ state size
stamp_dir <- fs::path(demo_dir, ".stamp")
info <- fs::dir_info(stamp_dir, recurse = TRUE)
data.table::data.table(stamp_state_mb = sum(info$size) / 1024^2)

# Find artifacts with large version histories
# Version directories are now distributed: <artifact-dir>/versions/
all_files <- fs::dir_info(demo_dir, recurse = TRUE, all = TRUE)
version_files <- all_files[grepl("/versions/", all_files$path), ]
if (nrow(version_files) > 0) {
  # Group by artifact and sum sizes
  version_files$artifact <- dirname(dirname(version_files$path))
  agg <- aggregate(size ~ artifact, version_files, sum)
  agg[order(-agg$size), ][1:10, ]
}
```

**Solution:**
```{r fix-disk, eval=FALSE}
# Configure retention policy to auto-prune old versions
st_opts(retention_policy = list(n = 10, days = 90))

# Manually prune versions for specific artifact
st_prune_versions(test_path, policy = list(n = 5), dry_run = FALSE)

# Prune all artifacts (use with caution!)
catalog <- stamp:::.st_catalog_read()
for (aid in unique(catalog$versions$artifact_id)) {
  artifact_path <- catalog$artifacts[artifact_id == aid]$path[1]
  st_prune_versions(artifact_path, policy = list(n = 5), dry_run = FALSE)
}
```

### 6.4 Version Timestamp Issues

**Symptom:** Timestamps appear corrupted or versions won't load.

**Diagnosis:**
```{r diagnose-timestamps}
# Check for invalid timestamps
vers <- st_versions(test_path)
bad_timestamps <- vers[is.na(created_at) | created_at == ""]
nrow(bad_timestamps)
```

**Solution:**
The package automatically drops corrupt version rows when reading. If this happens frequently, check for:

- System clock issues
- Concurrent access without proper locking
- Manual modification of catalog files

### 6.5 Lock File Issues

**Symptom:** `catalog.lock` file persists after crash.

**Solution:**
```{r fix-lock, eval=FALSE}
# Safe to delete if no stamp operations are running
lock_file <- fs::path(stamp_dir, "catalog.lock")
if (fs::file_exists(lock_file)) {
  fs::file_delete(lock_file)
}
```

---

## 7. Best Practices

### 7.1 Version Control Integration

**Include in `.gitignore`:**
```
.stamp/temp/
.stamp/logs/
.stamp/catalog.lock
```

**Consider including:**
- `.stamp/catalog.qs2` - Enables version history tracking across team
- `*/versions/` - Version snapshots for all artifacts (useful for small datasets)

**For large projects:**
```
# .gitignore
.stamp/catalog.qs2  # Local catalog only
*/versions/         # Exclude all version history (too large for git)
*/*.qs2             # Optionally exclude large binary artifacts
*/*.rds
```

**Note:** With the new distributed structure, version directories are scattered throughout your project (e.g., `data/model.rds/versions/`, `results/output.qs2/versions/`). Use wildcard patterns to exclude them.

### 7.2 Backup Strategy

```{r backup-strategy, eval=FALSE}
# Periodic catalog backup
backup_dir <- fs::path(demo_dir, "_backups")
fs::dir_create(backup_dir)

# Backup the catalog (small, fast)
catalog_src <- fs::path(demo_dir, ".stamp", "catalog.qs2")
catalog_dst <- fs::path(backup_dir, sprintf("catalog_%s.qs2", Sys.Date()))
fs::file_copy(catalog_src, catalog_dst, overwrite = TRUE)

# For critical projects, backup specific artifacts with version history
# Version directories are now distributed: <artifact-dir>/versions/
artifact_dir <- fs::path(demo_dir, "results", "model.rds")
if (fs::dir_exists(artifact_dir)) {
  artifact_backup <- fs::path(backup_dir, sprintf("model_artifact_%s", Sys.Date()))
  fs::dir_copy(artifact_dir, artifact_backup, overwrite = TRUE)
}

# Or backup the entire project directory (includes all artifacts + versions)
project_backup <- fs::path(backup_dir, sprintf("project_%s", Sys.Date()))
fs::dir_copy(demo_dir, project_backup, overwrite = TRUE)
```



---

## 8. Summary

The `stamp` package uses a **simplified, distributed storage architecture**:

✅ **Transparent storage** - Artifacts live at `<path>/<filename>/<filename>`, matching your mental model  
✅ **Distributed versioning** - Each artifact has its own `versions/` directory, not centralized  
✅ **State separation** - `.stamp/` contains only catalog, locks, and temp files  
✅ **Complete version history** - Every save creates an immutable snapshot next to the artifact  
✅ **Lineage tracking** - Parent relationships preserved in `parents.json`  
✅ **Concurrent access** - File-based locking prevents corruption  
✅ **Retention policies** - Auto-prune old versions per-artifact to manage disk space  

**Key takeaways for users:**

- Artifacts are stored as `<root>/<path>/<filename>/<filename>` (direct-path model)
- Version history lives in `<artifact-dir>/versions/`, not `.stamp/versions/`
- `.stamp/` contains only state (catalog, temp, logs) - no artifact data
- Use `st_versions()`, `st_info()`, and `st_load(version=...)` to explore history
- Safe to re-initialize without losing history

**Key takeaways for developers:**

- Catalog is the source of truth (two tables: artifacts, versions)
- Version snapshots are per-artifact and immutable
- All writes use atomic operations + locking for safety
- `.st_versions_root()` is deprecated (removed centralized version storage in v0.0.9)
- Internal functions prefixed with `.st_` provide infrastructure

---

```{r cleanup, include=FALSE}
# Clean up demo directory
fs::dir_delete(demo_dir)
```

## Further Reading

- `vignette("setup-and-basics", package = "stamp")` - General introduction
- `vignette("hashing-and-versions", package = "stamp")` - Deep dive into content hashing
